---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Welcome

<!-- badges: start -->
<!-- badges: end -->

The `envx` package aims to be a one-stop package for ENVX1002 and ENVX2001 students at the University of Sydney. It installs all the packages you need for the course, and provides a set of functions to help you with your assignments and tutorials.

**As of this writing, the package is still under development and contains only a few tutorials in development.**

[Contributing notes for the development team.](CONTRIBUTING.md)

## Installation

You can install the development version of envx from [GitHub](https://github.com/) with:

``` r
install.packages("remotes")
remotes::install_github("ENVX-resources/envx")
```


## Tutorials

The `envx` package includes a set of tutorials which accompany your labs. You can list the available tutorials with the `list_tutorials()` function:


```{r example}
library(envx)
list_tutorials()
```

When prompted, return `y` and then the number of the tutorial you would like to open.

