---
title: Two-sample t-tests
author: 
  - ---
  - "**ENVX1002 Introduction to Statistical Methods**"
  - School of Life and Environmental Science
  - The University of Sydney
  - ---
output: 
  learnr::tutorial:
    theme: cosmo
    language: 
      en:
        button:
          nexttopic: Next Page
          previoustopic: Previous Page
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, eval=TRUE, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)

cattle <- readRDS("cattle.rds")
```

## Introduction

![](images/usyd_horizontal.svg)

Welcome! In this tutorial you will analyse some data using the **two-sample *t*-test**. This test is used to determine if two groups have different means.

Work through the tutorial by answering the questions and completing the exercises. If you get stuck, there are hints and solutions available to help you. You will not be able to save your progress, so make sure you complete the tutorial in one sitting and copy code you want to keep for later. However, you can always come back and start again if you need to.

### Learning outcomes

- Recognise when to use a two-sample *t*-test.
- Use R and `ggplot2` to visualise data meant for a two-sample *t*-test.
- Perform a two-sample *t*-test.
- Understand the assumptions of a two-sample *t*-test.
- Interpret the results of a two-sample *t*-test.

## Cattle

### Background

Two breeds of cattle (brown and black) were investigated to see if one had a higher weight on average than the other when fed on the same diet. Farmers may use this information to decide which breed to raise for meat production.

### Data

The data is stored in the `cattle` data frame. You may view the data by running (*we also call it "printing", even though the `print()` function isn't used*) the data frame in the console.

Click on **Run Code** or use the shortcut `Ctrl + Enter` to run the code in the code chunk.

```{r ex-print, exercise=TRUE}
cattle
```

### Data exploration

It is important to know how to recognise, and visualise data that is suitable for a two-sample *t*-test. Let's start by checking the structure of the data. Which function would you use? Select **Hints** if you need help. 

```{r ex-str, exercise=TRUE}

```

```{r ex-str-hint}
# This function returns the structure of the data frame and
# displays the number of observations and variables.
```

```{r ex-str-solution}
str(cattle)
```


### 
Looks a bit messy, doesn't it! The `str()` function also stores attributes of the data frame, which is why it looks a bit cluttered. 

### 
We can use the `glimpse()` function from the `dplyr` package to get a tidier view of the data.

```{r}
dplyr::glimpse(cattle)
```


### 
When we're looking at the structure of the data, we're not only interested in the number of observations and variables, but also the type of variables. In this case, we have two variables: `Weight` and `Breed`. 

### 
The `Weight` variable is **continuous**, while the `Breed` variable is **categorical**. However, `Breed` is currently categorised as a character variable. 

```{r}
class(cattle$Breed)
```

We need to convert it to a factor variable.

### Wait! Why would we convert it to a factor variable?
If we did not collect the data ourselves, we may not know the type of variable `Breed` is. To determine the type of variable, we can:

- get context from the data source, or
- check the data dictionary that comes with the data, or
- check if it has a limited number of unique values by using the `unique()` function.

### 

If we cannot determine the type of variable from the data source or the data dictionary, we can use the `unique()` function to check the number of unique values. If the number of unique values is small, then it is likely a factor variable. Use the `unique()` function to check the number of unique values in the `Breed` variable.

```{r ex-factor, exercise=TRUE}

```

```{r ex-factor-hint}
# Your need to specifically select the Breed variable within 
# the cattle object and use the unique() function.
```

```{r ex-factor-solution}
unique(cattle$Breed)
```

###
Out of the 27 rows, the `Breed` variable has two unique values: `Brown` and `Black`. This means that it is a factor variable. So we can convert it to a factor variable.

### Converting to a factor variable
Below are two methods to do so. The first method uses base R, while the second method uses the `tidyverse` package. Which method you use is up to you, but the `tidyverse` method is (eventually) more readable and easier to understand due to the use of [pipes](https://style.tidyverse.org/pipes.html).


### {.tabset}
#### Base R method
```{r ex-factor-base, exercise=TRUE}
cattle$Breed <- as.factor(cattle$Breed)
```

#### Using `tidyverse`
```{r ex-factor-tidyverse, exercise=TRUE}
library(tidyverse)

# Using the cattle data frame, mutate the Breed variable 
# to a factor variable
cattle <- cattle %>% 
  mutate(Breed = as.factor(Breed))
```

### 
We can check the structure of the data again to confirm that the `Breed` variable has been converted to a factor variable. We'll just do it now for you so you don't have to run the code. Instead of `str()` let's use `glimpse()` from the `dplyr` package.

```{r include=FALSE}
cattle$Breed <- as.factor(cattle$Breed)
```

```{r}
dplyr::glimpse(cattle)
```

### 
It might be tempting to jump straight into the two-sample *t*-test, **but it is important to visualise the data first**. This will help us to understand the distribution of the data and identify any potential issues even *before* we perform our "official" assumptions checks. We will do this in the next section.

### 

```{r quiz1, echo=FALSE}
quiz(caption = "Test yourself before we move on", 
  ## QUESTION 1
  question("In the `cattle` data which is the response variable?",
    answer("Weight", correct = TRUE, message = "Correct! The response variable is the variable we are trying to predict."),
    answer("Breed"),
    incorrect = "Incorrect. The response variable is the variable we are trying to predict."
  ),
  ## QUESTION 2
  question("Intuitively, why is it important to check the structure of the data?",
    answer("To understand the type of variables in the data.", 
      correct = TRUE, 
      message = "Correct! Knowing the **type** of variables in the data helps us to choose the appropriate statistical test since the *t*-test requires at least one continuous variable and the other variable to be categorical (with two levels)."),
    answer("To know the number of variables and observations in the data.", 
      correct = TRUE, 
      message = "Knowing the **number** of variables in the data helps us think about how to visualise the data and whether we have enough data to perform the analysis."),
    answer("To identify whether the data requires some processing before analysis.", 
      correct = TRUE, 
      message = "And sometmes, data requires **processing** before analysis, such as conversion from character to factor variables and identification of missing values."),
    incorrect = "Perhaps you should think some more (hint: think about the words in **bold**.",
    allow_retry = TRUE
  )
)
```



## Visualising two groups

The **boxplot** is a good choice when comparing two groups of data and is often the first plot to create when visualising data for a two-sample *t*-test. It allows us to compare between groups while also retaining information about the distribution of the data, which is important for the test.

###
In this tutorial we will create a boxplot that is **fit for scientific publication**. Usually this is the *one plot* that you will decide to use to graphically represent your data. Often called a **summary plot**, it should be clear, concise, and informative as it summarises the data in a single plot.

###
This is in contrast to **exploratory plots** and **diagnostic plots**, which are used to explore the data, check assumptions, and identify issues. Such plots are not usually included in scientific publications but are important for the researcher to understand the data.


### Using `ggplot2` to create a boxplot

Load the `ggplot2` package.


```{r}
library(ggplot2)
```

This package is also loaded if you are using the `tidyverse` package, so `library(tidyverse)` will also load `ggplot2`.

###
Plotting in `ggplot2` requires data to be in a *specific* format. This format is often referred to as "long" format, and in data science circles it is also called "tidy" data. The `cattle` data is already in this format -- each **row is an observation** and each **column is a variable**. 

If your data is not in this format e.g. if you have a column for each breed, you will need to manipulate the data to be in this format. You can do this reproducibly in R using the `tidyverse` package, or manually in Excel or Google Sheets.

### 
The advantages of "tidy" data are [numerous](https://vita.had.co.nz/papers/tidy-data.pdf), but the most important one is that R functions are designed to work with data in this format -- so you will save time and effort by keeping your data tidy.

Anyway, let's plot the data.

### 

The most basic plot can be created using the code below:

```{r}
ggplot(data = cattle, mapping = aes(x = Breed, y = Weight)) +
  geom_boxplot()
```

###

Above, we are simply calling the `ggplot()` function and specifying the data and aesthetics. We are then adding a `geom_boxplot()` layer to create the boxplot.

It works, but it will **not** pass muster for a scientific publication. We need to add labels, a title, and a figure caption. We also want to (in most cases) reduce distractions by removing the gridlines and background colour.

###
Labels can be added by adding more "layers" to the plot using the `+` operator. The `labs()` function is used to add labels, titles, and captions. We will only add labels for the *x-axis* and *y-axis*. Titles are not necessary in most cases unless we have a multi-panel figure.

```{r}
ggplot(data = cattle, mapping = aes(x = Breed, y = Weight)) +
  geom_boxplot() +
  labs(
    x = "Breed",
    y = "Weight (kg)",
  )
```

###
To remove the grid lines and background colour, we can either muck around with the theme settings or use the `theme_classic()` function. This function is part of the `ggplot2` package and is used to remove the background colour and gridlines. 

```{r}
ggplot(data = cattle, mapping = aes(x = Breed, y = Weight)) +
  geom_boxplot() +
  labs(
    x = "Breed",
    y = "Weight (kg)",
  ) +
  theme_classic()
```

###
Figures need captions, and applying them uses different methods depending on whether you are producing a `.Rmd` document or a Quarto `.qmd` document. Assuming that you are using Quarto, you want to use the `label` and `fig-cap` options in the code chunk.

````{verbatim}
```{r}
#| label: fig-cattle-boxplot
#| fig-cap: "Boxplot of cattle weight (kg) by breed."

ggplot(data = cattle, mapping = aes(x = Breed, y = Weight)) +
  geom_boxplot() +
  labs(
    x = "Breed",
    y = "Weight (kg)",
  ) +
  theme_classic()
```
````

This will show the figure caption below the plot. Note that execution options are not visible in the final document (so your `#|` lines will not be visible).

```{r fig.cap="Figure 1. Boxplot of cattle weight (kg) by breed."}
ggplot(data = cattle, mapping = aes(x = Breed, y = Weight)) +
  geom_boxplot() +
  labs(
    x = "Breed",
    y = "Weight (kg)",
  ) +
  theme_classic()
```

###
The plot is now ready for publication, but we have provided the **minimal** amount of information. You can add more layers to the plot to make it more informative. For example, you can change the colours, flip the axes, or add a title. To do so it is important to understand how to search for help using Google, or prompt GenAI to help you, which we will not cover in this tutorial.

###
You can now go through the code above and run it in this document below. Play around with the code to see how it changes the plot. Once you are ready, hit the **Continue** (or **Next Page**, depending on how you navigated this tutorial) button to move on.

```{r ex-boxplot, exercise=TRUE}
# Add the code to create the boxplot here
```


## HATPC

Let' use the **HATPC** method to guide us through the statistical analysis for a two-sample *t*-test. 
The steps are:

1. **Hypothesis**: State the null and alternative hypotheses.
2. **Assumptions**: State the assumptions of the test.
3. **Test**: Perform the test (calculate the test statistic).
4. **P-value**: Calculate the p-value.
5. **Conclusion**: Make a conclusion based on the p-value.

###
Work through these steps should be routine for you by now. If you are unsure about any of the steps, you can always refer back to the previous labs and tutorials.

### Hypothesis

```{r q-hypothesis, echo=FALSE}
quiz(caption = "Null and alternate hypotheses",
  question("What is the null hypothesis for the two-sample t-test?",
    answer("The two groups have the same mean weight.", correct = TRUE),
    answer("The two groups have different mean weights."),
    answer("The two groups have the same median weight."),
    answer("The two groups have different median weights."),
    random_answer_order = TRUE),
  question("What is the alternative hypothesis for the two-sample t-test?",
    answer("The two groups have different mean weights.", correct = TRUE),
    answer("The two groups have the same mean weight."),
    answer("The 'Brown' group has a higher mean weight than the 'Black' group."),
    answer("The 'Black' group has a higher mean weight than the 'Brown' group."),
    random_answer_order = TRUE)
)
```

### Assumptions

The assumptions of the two-sample *t*-test are:

1. **Normality**: The data in each group is normally distributed.
2. **Equal variance**: The variance in each group is equal.

### Normality

Lecture 6a covered the normality assumption. Try to work out how you would do it for this data set. 

```{r ex-normality, exercise=TRUE}
# Click on Hints for help (and answers).
```

```{r ex-normality-hint-1}
# You can use the boxplot, histogram, or qqplot 
# to check for normality. You can also use a formal
# statistical test such as the Shapiro-Wilk test.
```


```{r ex-normality-hint-2}
# Using a boxplot should be simple as you have already done so
# in the previous exercise.
ggplot(data = cattle, mapping = aes(x = Weight)) +
  geom_histogram()
```

```{r ex-normality-hint-3}
# QQ-plot
ggplot(data = cattle, mapping = aes(sample = Weight)) +
  stat_qq() +
  stat_qq_line()
```

```{r ex-normality-hint-4}
# Histogram
ggplot(data = cattle, mapping = aes(x = Weight)) +
  geom_histogram() +
  facet_wrap(~ Breed)
```

```{r ex-normality-hint-5}
# Shapiro-Wilk test
shapiro.test(cattle$Weight)
```



### Equal variances

Again, use Lecture 6a to work out how you would check for equal variances.

```{r ex-equal-variance, exercise=TRUE}
# Click on Hints for help (and answers).
```

```{r ex-equal-variance-hint-1}
# You can use the boxplot to check for equal variances, or
# perform a statistical test.
```

```{r ex-equal-variance-hint-2}
# Levene's test
library(car)
leveneTest(Weight ~ Breed, data = cattle)
```


```{r ex-equal-variance-hint-3}
# Bartlett's test
bartlett.test(Weight ~ Breed, data = cattle)
```

###
```{r q-assumptions, echo=FALSE}
quiz(caption = "Assumption checks",
  question("Did the data meet the assumptions of normality?",
    answer("Yes", correct = TRUE),
    answer("No"),
    incorrect = "You may need to check the plots and statistical tests again, or refer back to the lecture notes. This is an important step in the analysis."
  ),
  question("Did the data meet the assumptions of equal variances?",
    answer("Yes", correct = TRUE),
    answer("No"),
    incorrect = "You may need to check the plots and statistical tests again, or refer back to the lecture notes. This is an important step in the analysis."
  )
)
```

### Test

Run the two-sample *t*-test using the `t.test()` function in R. **Remember to specify whether the equal variance assumption is met or not in the function!**

```{r ex-t-test, exercise=TRUE}
# Run the two-sample t-test
```

```{r ex-t-test-hint}
# You need to specify the formula and the data frame.
```

```{r ex-t-test-solution}
fit <- t.test(Weight ~ Breed, data = cattle, var.equal = TRUE)
fit
```


### P-value and conclusion

Let's view the results again:

```{r}
fit <- t.test(Weight ~ Breed, data = cattle, var.equal = TRUE)
fit
```

You should know how to interpret the results by now. If you are unsure, refer back to the lecture notes or the previous tutorials.

You may write your conclusion in the box below. Click **Continue** to view solutions.

```{r ex-conclusion, exercise=TRUE}
# Write your conclusion here using the comment tag `#`
```

### 

#### Solution

We check the assumptions graphically by plotting both a QQ-plot and a boxplot of the data.

```{r echo=FALSE}
# Normality
ggplot(data = cattle, mapping = aes(sample = Weight)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~ Breed)

# Equal variances
ggplot(data = cattle, mapping = aes(x = Breed, y = Weight)) +
  geom_boxplot()
```

Both assumptions were met:

- The data is normally distributed for both groups, as shown by the QQ-plot and boxplot.
- The equal variances assumption is met, as the range of data shown between the two boxplots is not orders of magnitude different.

**Note**: because assumption checks are exploratory, we do not need to report the results in detail, neither do the plots need to "look pretty" or be publication-ready. They are for our own understanding.

**Interpretation and conclusion:** The results of the two-sample t-test show that the p-value is less than 0.05. This means that we reject the null hypothesis and conclude that the two groups have different mean weights. From the results, Brown cows have a significantly higher mean weight than Black cows.




## Thanks

You have reached the end of this tutorial. We hope you have learned how to perform a two-sample *t*-test and interpret the results.


*This document is developed using resources that are available under a [Creative Commons Attribution 4.0 International license](https://github.com/usyd-soles-edu/.github/blob/main/cc-by), made available on the [SOLES Open Educational Resources repository](https://github.com/usyd-soles-edu).*